Robinson et al. 2020 Supplemental Information: Comparing the Radiocarbon and Tree-ring Records for the Upland US Southwest

### 2020-01-19 analysis/paper/paper.Rmd
uploaded more memory efficient version

---
title: "Robinson et al. 2020 Supplemental Information: Comparing the Radiocarbon and Tree-ring Records for the Upland US Southwest"
author:
  - Kyle Bocinsky:
      email: bocinsky@gmail.com
      institute: [CCAC]
      correspondence: true
  - Erick Robinson:
      email: erick.robinson@usu.edu
      institute: [USU]
      correspondence: false
institute:
  - CCAC: Crow Canyon Archaeological Center
  - USU: Utah State University
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
      code_folding: hide
      fig_caption: yes
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/philosophical-transactions-of-the-royal-society-b.csl" # Insert path for the bib-style
editor_options: 
  chunk_output_type: console
---

@@

```{r prepare-radiocarbon-data}
# # Read in the radiocarbon database
# # radiocarbon_data <-
#   readr::read_csv(here::here("analysis/data/raw_data/0403_sbox_clean.csv")) %>%
#   # Convert to spatial object
#   sf::st_as_sf(coords = c("Long","Lat"),
#                crs = 4326) %>%
#   sf::st_transform(four_corners_lcc_proj) %>%
#   # Select salient columns
#   dplyr::select(labnumber,
#                 date,
#                 sd,
#                 Site,
#                 SiteID) %>%
#   dplyr::rename(Lab_Number = labnumber,
#                 Date_BP = date,
#                 SD = sd,
#                 Site_Number = Site) %>%
#   sf::st_intersection(uusw_counties) %>%
#   sf::st_drop_geometry() %>%
#   dplyr::mutate(Type = "Radiocarbon",
#                 SiteID = SiteID %>% stringr::str_c("Radiocarbon - ", .)) %>%
#   dplyr::select(Type,
#                 SiteID,
#                 Site_Number,
#                 County,
#                 Lab_Number,
#                 Date_BP,
#                 SD) %T>%
#   # Write the output to a new file (csv)
#   readr::write_csv(here::here("analysis/data/derived_data/radiocarbon_data.csv"))

radiocarbon_data <-
  readr::read_csv(here::here("analysis/data/derived_data/radiocarbon_data.csv"))

@@

We draw on the tree-ring date database presented in @Bocinsky2016_SA and archived in the Digital Archaeological Record (https://doi.org/10.6067/XCV82J6D7B). The tree-ring data we use here were compiled by Tim Kohler and Rebecca Higgins from many contributers, and the vast majority of dates were determined by researchers in the Laboratory of Tree Ring Research at the University of Arizona. The original dataset contained 32,863 cutting, near-cutting, and non-cutting dates from across the US Southwest. Here, we only use the dates that are most likely to be accurate --- cutting dates ('B', 'G', 'L', 'c', or 'r' dates, with or without a '+') for which we know the year the tree died, and near-cutting dates ('v' or 'v+' dates) which are likely within 0--3 years of the true date of the outermost ring. The @Bocinsky2016_SA data contain site locations; for this analysis, we first spatially cropped the database to our study area, and then translated those locations to the county level. We converted the original date determinations to BP to match radiocarbon conventions.

```{r prepare-treering-data}
# # Read in the Bocinsky et al. 2016 tree ring data
# # https://doi.org/10.6067/XCV82J6D7B
# dendro_data <-
#   readr::read_csv(here::here("analysis/data/raw_data/BOCINSKY_ET_AL_2015_TREE_RINGS.csv")) %>%
#   # Convert to spatial object
#   sf::st_as_sf(coords = c("LON","LAT"),
#                crs = 4326) %>%
#   sf::st_transform(four_corners_lcc_proj) %>%
#   # Convert dates to BP
#   dplyr::mutate(Date_BP = 1950 - Outer_Date_AD,
#                 Type = ifelse(C_Level %in% c(2,3),
#                               "Cutting or Near-cutting",
#                               "Non-cutting")) %>%
#   # drop C_Level
#   dplyr::select(Seq_Number,
#                 Site_Number,
#                 Lab_Number,
#                 Type,
#                 Date_BP) %>%
#   dplyr::rename(SiteID = Seq_Number) %>%
#   sf::st_intersection(uusw_counties) %>%
#   sf::st_drop_geometry() %>%
#   dplyr::mutate(Type = Type %>% stringr::str_c("Dendro - ", .),
#                 SiteID = SiteID %>% stringr::str_c("Dendro - ", .),
#                 SD = 0) %>%
#   dplyr::filter(Type == "Dendro - Cutting or Near-cutting") %>%
#   dplyr::mutate(Type = "Tree-ring")  %>%
#   dplyr::select(Type,
#                 SiteID,
#                 Site_Number,
#                 County,
#                 Lab_Number,
#                 Date_BP,
#                 SD) %T>%
#   # Write the output to a new file (csv)
#   readr::write_csv(here::here("analysis/data/derived_data/dendro_data.csv"))

dendro_data <- 
  readr::read_csv(here::here("analysis/data/derived_data/dendro_data.csv"))

@@

```{r data-summary}
all_data %>%
  dplyr::bind_rows() %>%
  dplyr::bind_rows(all_data %>%
                     dplyr::bind_rows() %>%
                     dplyr::mutate(`Study Area` = "UUSW Total")) %>%
  dplyr::filter(!is.na(`Study Area`)) %>%
  dplyr::mutate(`Study Area` = factor(`Study Area`, levels = c("UUSW Total","CMV","NRG"))) %>%
  dplyr::group_by(Type,`Study Area`) %>%
  dplyr::summarise(Count = dplyr::n()) %>%
  tidyr::spread(Type, Count) %>%
  knitr::kable(caption = "Counts of radiocarbon and tree-ring dates in this study.",
               booktabs = TRUE,
               format.args = list(big.mark = ','))
```

@@

if(!file.exists(here::here("analysis/data/derived_data/radiocarbon_spd.rds")))
  all_data$Radiocarbon %>%
  split(.,.$`Study Area`) %>%
  tibble::tibble(`Study Area` = names(.),
                 Dates = .) %>%
  dplyr::bind_rows(tibble::tibble(`Study Area` = "UUSW",
                                  Dates = list(all_data$Radiocarbon))) %>%
  dplyr::rowwise() %>%
  dplyr::mutate(Calibrated = list(Dates %$%
                                    rcarbon::calibrate(x = Date_BP,
                                                       errors = SD,
                                                       timerange = timerange,
                                                       calMatrix = TRUE,
                                                       verbose = FALSE)),
                Bins = list(rcarbon::binPrep(sites = Dates$SiteID,
                                             ages = Calibrated,
                                             h = 100)),
                SPD = list(rcarbon::spd(x = Calibrated,
                                        bins = Bins,
                                        datenormalised = TRUE,
                                        spdnormalised = TRUE,
                                        timeRange = c(1949,0),
                                        runm = 21,
                                        verbose = FALSE))) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(Density = SPD %>%
                  purrr::map(function(x){
                    x$grid %>%
                      tibble::as_tibble() %$%
                      tibble::tibble(Type = "Radiocarbon Density",
                                     Date_BP = calBP,
                                     Density = PrDens / sum(PrDens, na.rm = T))
                  })
  ) %>%
  readr::write_rds(here::here("analysis/data/derived_data/radiocarbon_spd.rds"),
                   compress = "gz")

radiocarbon_spd <- readr::read_rds(here::here("analysis/data/derived_data/radiocarbon_spd.rds"))

@@

if(!file.exists(here::here("analysis/data/derived_data/dendro_spd.rds")))
  all_data$`Tree-ring` %>%
  split(.,.$`Study Area`) %>%
  tibble::tibble(`Study Area` = names(.),
                 Dates = .) %>%
  dplyr::bind_rows(tibble::tibble(`Study Area` = "UUSW",
                                  Dates = list(all_data$`Tree-ring`))) %>%
  dplyr::rowwise() %>%
  dplyr::mutate(Calibrated = list(Dates %$%
                                    calibrate_tree_ring_dates(x = Date_BP)),
                Bins = list(rcarbon::binPrep(sites = Dates$SiteID,
                                             ages = Dates$Date_BP,
                                             h = 100)),
                SPD = list(rcarbon::spd(x = Calibrated,
                                        bins = Bins,
                                        datenormalised = TRUE,
                                        spdnormalised = TRUE,
                                        timeRange = timerange,
                                        runm = 21,
                                        verbose = FALSE))) %>%

  dplyr::ungroup() %>%
  dplyr::mutate(Density = SPD %>%
                  purrr::map(function(x){
                    x$grid %>%
                      tibble::as_tibble() %$%
                      tibble::tibble(Type = "Tree-ring Density",
                                     Date_BP = calBP,
                                     Density = PrDens / sum(PrDens, na.rm = T))
                  })
  ) %>%
  readr::write_rds(here::here("analysis/data/derived_data/dendro_spd.rds"),
                   compress = "gz")

dendro_spd <- readr::read_rds(here::here("analysis/data/derived_data/dendro_spd.rds"))

>>>>

---
title: "Robinson et al. 2020 Supplemental Information: Comparing the Radiocarbon and Tree-ring Records for the Upland US Southwest"
author:
  - Kyle Bocinsky:
      email: bocinsky@gmail.com
      institute: [CCAC]
      correspondence: true
  - Erick Robinson:
      email: erick.robinson@usu.edu
      institute: [USU]
      correspondence: false
institute:
  - CCAC: Crow Canyon Archaeological Center
  - USU: Utah State University
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
      code_folding: hide
      fig_caption: yes
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::word_document2:
      reference_docx: "../templates/template.docx"
      code_folding: hide
      fig_caption: yes
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/philosophical-transactions-of-the-royal-society-b.csl" # Insert path for the bib-style
editor_options: 
  chunk_output_type: console
---

@@

```{r prepare-radiocarbon-data}
# # Read in the radiocarbon database
# # radiocarbon_data <-
# readr::read_csv(
#   here::here("analysis/data/raw_data/0403_sbox_clean.csv")
# ) %>%
#   # Convert to spatial object
#   sf::st_as_sf(coords = c("Long","Lat"),
#                crs = 4326) %>%
#   sf::st_transform(four_corners_lcc_proj) %>%
#   # Select salient columns
#   dplyr::select(labnumber,
#                 date,
#                 sd,
#                 Site,
#                 SiteID) %>%
#   dplyr::rename(Lab_Number = labnumber,
#                 Date_BP = date,
#                 SD = sd,
#                 Site_Number = Site) %>%
#   sf::st_intersection(uusw_counties) %>%
#   sf::st_drop_geometry() %>%
#   dplyr::mutate(Type = "Radiocarbon",
#                 SiteID = SiteID %>% stringr::str_c("Radiocarbon - ", .)) %>%
#   dplyr::select(Type,
#                 SiteID,
#                 Site_Number,
#                 County,
#                 Lab_Number,
#                 Date_BP,
#                 SD) %T>%
#   # Write the output to a new file (csv)
#   readr::write_csv(
#     here::here("analysis/data/derived_data/radiocarbon_data.csv")
#   )

radiocarbon_data <-
  readr::read_csv(
    here::here("analysis/data/derived_data/radiocarbon_data.csv")
  )

@@

We draw on the tree-ring date database presented in @Bocinsky2016_SA and archived in the Digital Archaeological Record (https://doi.org/10.6067/XCV82J6D7B). The tree-ring data we use here were compiled by Tim Kohler and Rebecca Higgins from many contributors, and the vast majority of dates were determined by researchers in the Laboratory of Tree Ring Research at the University of Arizona. The original dataset contained 32,863 cutting, near-cutting, and non-cutting dates from across the US Southwest. Here, we only use the dates that are most likely to be accurate --- cutting dates ('B', 'G', 'L', 'c', or 'r' dates, with or without a '+') for which we know the year the tree died, and near-cutting dates ('v' or 'v+' dates) which are likely within 0--3 years of the true date of the outermost ring. The @Bocinsky2016_SA data contain site locations; for this analysis, we first spatially cropped the database to our study area, and then translated those locations to the county level. We converted the original date determinations to BP to match radiocarbon conventions.

```{r prepare-treering-data}
# # Read in the Bocinsky et al. 2016 tree ring data
# # https://doi.org/10.6067/XCV82J6D7B
# dendro_data <-
#   readr::read_csv(
#     here::here("analysis/data/raw_data/BOCINSKY_ET_AL_2015_TREE_RINGS.csv")
#   ) %>%
#   # Convert to spatial object
#   sf::st_as_sf(coords = c("LON","LAT"),
#                crs = 4326) %>%
#   sf::st_transform(four_corners_lcc_proj) %>%
#   # Convert dates to BP
#   dplyr::mutate(Date_BP = 1950 - Outer_Date_AD,
#                 Type = ifelse(C_Level %in% c(2,3),
#                               "Cutting or Near-cutting",
#                               "Non-cutting")) %>%
#   # drop C_Level
#   dplyr::select(Seq_Number,
#                 Site_Number,
#                 Lab_Number,
#                 Type,
#                 Date_BP) %>%
#   dplyr::rename(SiteID = Seq_Number) %>%
#   sf::st_intersection(uusw_counties) %>%
#   sf::st_drop_geometry() %>%
#   dplyr::mutate(Type = Type %>% stringr::str_c("Dendro - ", .),
#                 SiteID = SiteID %>% stringr::str_c("Dendro - ", .),
#                 SD = 0) %>%
#   dplyr::filter(Type == "Dendro - Cutting or Near-cutting") %>%
#   dplyr::mutate(Type = "Tree-ring")  %>%
#   dplyr::select(Type,
#                 SiteID,
#                 Site_Number,
#                 County,
#                 Lab_Number,
#                 Date_BP,
#                 SD) %T>%
#   # Write the output to a new file (csv)
#   readr::write_csv(
#     here::here("analysis/data/derived_data/dendro_data.csv")
#   )

dendro_data <- 
  readr::read_csv(
    here::here("analysis/data/derived_data/dendro_data.csv")
  )

@@

```{r data-summary}
all_data %>%
  dplyr::bind_rows() %>%
  dplyr::bind_rows(all_data %>%
                     dplyr::bind_rows() %>%
                     dplyr::mutate(`Study Area` = "UUSW Total")) %>%
  dplyr::filter(!is.na(`Study Area`)) %>%
  dplyr::mutate(`Study Area` = factor(`Study Area`, 
                                      levels = c("UUSW Total",
                                                 "CMV",
                                                 "NRG"))) %>%
  dplyr::group_by(Type,`Study Area`) %>%
  dplyr::summarise(Count = dplyr::n()) %>%
  tidyr::spread(Type, Count) %>%
  knitr::kable(caption = "Counts of radiocarbon and tree-ring dates in this study.",
               booktabs = TRUE,
               format.args = list(big.mark = ','))
```

@@

if(!file.exists(
  here::here("analysis/data/derived_data/radiocarbon_spd.rds")
))
  all_data$Radiocarbon %>%
  split(.,.$`Study Area`) %>%
  tibble::tibble(`Study Area` = names(.),
                 Dates = .) %>%
  dplyr::bind_rows(tibble::tibble(`Study Area` = "UUSW",
                                  Dates = list(all_data$Radiocarbon))) %>%
  dplyr::rowwise() %>%
  dplyr::mutate(Calibrated = list(Dates %$%
                                    rcarbon::calibrate(x = Date_BP,
                                                       errors = SD,
                                                       timerange = timerange,
                                                       calMatrix = TRUE,
                                                       verbose = FALSE)),
                Bins = list(rcarbon::binPrep(sites = Dates$SiteID,
                                             ages = Calibrated,
                                             h = 100)),
                SPD = list(rcarbon::spd(x = Calibrated,
                                        bins = Bins,
                                        datenormalised = TRUE,
                                        spdnormalised = TRUE,
                                        timeRange = c(1949,0),
                                        runm = 21,
                                        verbose = FALSE))) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(Density = SPD %>%
                  purrr::map(function(x){
                    x$grid %>%
                      tibble::as_tibble() %$%
                      tibble::tibble(Type = "Radiocarbon Density",
                                     Date_BP = calBP,
                                     Density = PrDens / sum(PrDens, na.rm = T))
                  })
  ) %>%
  readr::write_rds(
    here::here("analysis/data/derived_data/radiocarbon_spd.rds"),
    compress = "gz")

radiocarbon_spd <- readr::read_rds(
  here::here("analysis/data/derived_data/radiocarbon_spd.rds")
)

@@

if(!file.exists(
  here::here("analysis/data/derived_data/dendro_spd.rds")
))
  all_data$`Tree-ring` %>%
  split(.,.$`Study Area`) %>%
  tibble::tibble(`Study Area` = names(.),
                 Dates = .) %>%
  dplyr::bind_rows(tibble::tibble(`Study Area` = "UUSW",
                                  Dates = list(all_data$`Tree-ring`))) %>%
  dplyr::rowwise() %>%
  dplyr::mutate(Calibrated = list(Dates %$%
                                    calibrate_tree_ring_dates(x = Date_BP)),
                Bins = list(rcarbon::binPrep(sites = Dates$SiteID,
                                             ages = Dates$Date_BP,
                                             h = 100)),
                SPD = list(rcarbon::spd(x = Calibrated,
                                        bins = Bins,
                                        datenormalised = TRUE,
                                        spdnormalised = TRUE,
                                        timeRange = timerange,
                                        runm = 21,
                                        verbose = FALSE))) %>%

  dplyr::ungroup() %>%
  dplyr::mutate(Density = SPD %>%
                  purrr::map(function(x){
                    x$grid %>%
                      tibble::as_tibble() %$%
                      tibble::tibble(Type = "Tree-ring Density",
                                     Date_BP = calBP,
                                     Density = PrDens / sum(PrDens, na.rm = T))
                  })
  ) %>%
  readr::write_rds(
    here::here("analysis/data/derived_data/dendro_spd.rds"),
    compress = "gz")

dendro_spd <- readr::read_rds(
  here::here("analysis/data/derived_data/dendro_spd.rds")
)
